Brownian local times

### 2015-09-10 notes/brownian-local-times.Rmd
.

Let's check that, too: why not?
```{r killing, cache=TRUE, depends="sim_rw"}
loc <- 0.1
zeta <- function (x,r) {
    2*pnorm(abs(x)) - 1 + 2 * exp( r*abs(x) + r^2/2 + pnorm(abs(x)+r,lower.tail=FALSE,log.p=TRUE) )
}
compare.rates <- as.data.frame( t( sapply( c(0.25,0.5,1,1.5,2,2.5,3), function (rate) {
        visits <- colSums(abs(rws-loc)<=1.5*eps)
        killed <- ( rbinom(ncol(rws),size=visits,prob=rate/ntimes/(3*eps)) == 0 )
        c( 
          rate=rate,
          simulation=mean(killed), 
          theory=zeta(loc,rate),
          SE=sqrt(mean(killed)*(1-mean(killed))/nwalks)
          )
    } ) ) )
with(compare.rates, {
         plot( rate, theory, type='l', xlab="recomb rate", ylab="prob of recomb", col='red', ylim=range(c(simulation,theory)), ylim=c(0,1) ) 
         points( rate, simulation, pch=20 )
         arrows( x0=rate, y0=simulation-2*SE, y1=simulation+2*SE, angle=90, code=3 )
    } )
```

@@

The key quantity we want is the probability that a Brownian motion with variance $\sigma^2$ killed at rate $r$
with respect to its local time at $x$ was still alive at time $\tau$.
Denote this by
$$
\begin{align*}
    \zeta(x,r,\tau,\sigma).
\end{align*}
$$
By Brownian scaling,
since $B_{ct}$ has the same distribution as $\sqrt{c}B_t$,
$$
\begin{align*}
    \zeta(x,r,\tau,\sigma) 
    &=
    \zeta(x,r\tau,1,\sqrt{\tau}\sigma) \\
    &=
    \zeta(x/\sqrt{\tau\sigma^2},r\tau,1,1) ,
    % &=
    % 2 \exp\left(r\tau\left(\frac{|x|}{\sqrt{\tau}\sigma}+\frac{r\tau}{2}\right)\right) \left( 1-\Phi\left(\frac{|x|}{\sqrt{\tau\sigma^2}}+r\tau \right) \right) .
\end{align*}
$$
and
$$
\begin{align*}
    \zeta(x,r,1,1)
    &=
    2\Phi(|x|)-1 + 
    2 e^{r|x|+r^2/2} \left( 1-\Phi(|x|+r) \right) .
\end{align*}
$$

Now, let $p(x,r,\tau,A)$ denote the probability that a lineage begun at $x<0$
at a location at recombination distance $r$ from the selected site
inherits from the left-hand side of the barrier
(i.e., an ancestor on $(-\infty,0)$)
$\tau$ generations ago
is the probability no recombination at the barrier has occurred
plus one-half the probability recombination has occurred.
This is
$$
\begin{align*}
    p(x,r,\tau,A) = % \zeta 1_{x\le0} + (1-\zeta)/2 = (1-(1-2*1_{x\le0})\zeta)/2
    \frac{1}{2} \left\{ 1 - \sgn(x) \zeta(x,r,\tau) \right\} .
    % \frac{1}{2}\left\{ 1 - \sgn(x) 2 \exp\left(r\tau\left(\frac{x}{\sqrt{\tau}\sigma}+\frac{r\tau}{2}\right)\right) \left( 1-\Phi\left(\frac{x}{\sqrt{\tau\sigma^2}}+r\tau \right) \right) \right\}
\end{align*}
$$

Does this actually make a cline?
```{r show_cline}
pcline <- function (...) {
    (1/2)*( 1 - zeta(...) )
}
xx <- seq(-5,5,length.out=500)
plot(xx, pcline(xx,r=1), type='l' )
```

>>>>

Let's check that, too: why not?
```{r killing, cache=TRUE, depends="sim_rw"}
loc <- 0.1
zeta <- function (x,r) {
    2*pnorm(abs(x)) - 1 + 2 * exp( r*abs(x) + r^2/2 + pnorm(abs(x)+r,lower.tail=FALSE,log.p=TRUE) )
}
compare.rates <- as.data.frame( t( sapply( c(0.25,0.5,1,1.5,2,2.5,3), function (rate) {
        visits <- colSums(abs(rws-loc)<=1.5*eps)
        killed <- ( rbinom(ncol(rws),size=visits,prob=rate/ntimes/(3*eps)) == 0 )
        c( 
          rate=rate,
          simulation=mean(killed), 
          theory=zeta(loc,rate),
          SE=sqrt(mean(killed)*(1-mean(killed))/nwalks)
          )
    } ) ) )
with(compare.rates, {
         plot( rate, theory, type='l', xlab="recomb rate", ylab="prob of recomb", col='red', ylim=c(0,1) ) 
         points( rate, simulation, pch=20 )
         arrows( x0=rate, y0=simulation-2*SE, y1=simulation+2*SE, angle=90, code=3 )
    } )
```

This is very steep at $x=0$.
Notice that as $x$ goes to 0,
$$
\begin{align*}
    \lim_{x \to 0} \zeta(x,r) 
    &= 2 e^{r^2/2} (1-\Phi(r)) \\
    &= \frac{2}{r\sqrt{2\pi}} + O(r^{-3}) .
\end{align*}
$$
so that the difference in the left and right limits is
$$
\begin{align*}
    \lim_{x \to 0-} p(x,r) 
    -\lim_{x \to 0+} p(x,r) 
    &= 
    \frac{1}{2}+\frac{2}{r\sqrt{2\pi}} 
    - \frac{1}{2}-\frac{2}{r\sqrt{2\pi}} 
      +O(r^{-3}) \\
    &=
    \frac{4}{r\sqrt{2\pi}}
      +O(r^{-3}) 
\end{align*}
$$

Let's also look at the slope of $p$ at $x=0$.
Since $\partial_x |x| = \sgn(x)$,
$$
\begin{align*}
    \partial_x \zeta(x,r)
    &=
    2 \sgn(x) \left(
    \phi(|x|)
    + r e^{r|x|+r^2/2} (1-\Phi(|x|+r))
    - e^{r|x|+r^2/2} \phi(|x|+r)
      \right)
\end{align*}
$$
and so
$$
\begin{align*}
    -\partial_x p(x)
    &=
    \phi(|x|)
    + r e^{r|x|+r^2/2} (1-\Phi(|x|+r))
    - e^{r|x|+r^2/2} \phi(|x|+r) \\
    &\to
    \frac{1}{\sqrt{2\pi}}(1-e^{r^2/2}e^{-r^2/2})
    + r e^{r^2/2} (1-\Phi(r))  \qquad \text{as } x \to 0 \\
    &=
    r e^{r^2/2} (1-\Phi(r)) \\
    &=
    \frac{1}{\sqrt{2\pi}} + O(\frac{1}{r}) .
\end{align*}
$$

@@

The key quantity we want is the probability that a Brownian motion with variance $\sigma^2$ killed at rate $r$
with respect to its local time at $x$ was still alive at time $\tau$.
Denote this by
$$
\begin{align*}
    \zeta(x,r,\tau,\sigma).
\end{align*}
$$
By Brownian scaling,
since $B_{ct}$ has the same distribution as $\sqrt{c}B_t$,
$$
\begin{align*}
    \zeta(x,r,\tau,\sigma) 
    &=
    \zeta(x,r\tau,1,\sqrt{\tau}\sigma) \\
    &=
    \zeta(x/\sqrt{\tau\sigma^2},r\tau,1,1) ,
    % &=
    % 2 \exp\left(r\tau\left(\frac{|x|}{\sqrt{\tau}\sigma}+\frac{r\tau}{2}\right)\right) \left( 1-\Phi\left(\frac{|x|}{\sqrt{\tau\sigma^2}}+r\tau \right) \right) .
\end{align*}
$$
and
$$
\begin{align*}
    \zeta(x,r,1,1)
    &=
    1 - 2(1-\Phi(|x|)) + 
    2 e^{r|x|+r^2/2} \left( 1-\Phi(|x|+r) \right) .
\end{align*}
$$

Now, let $p(x,r,\tau,A)$ denote the probability that a lineage begun at $x<0$
at a location at recombination distance $r$ from the selected site
inherits from the left-hand side of the barrier
(i.e., an ancestor on $(-\infty,0)$)
$\tau$ generations ago.
For an individual to the left of the barrier ($x<0$),
this is the probability no recombination at the barrier has occurred
plus one-half the probability recombination has occurred,
i.e., $\zeta+(1-\zeta)/2=(1+\zeta)/2$.
If $x>0$ this is the one-half times probability a recombination has occurred,
i.e., $(1-\zeta)/2$.
Combining these,
the probability of ending up on the left-hand side is
$$
\begin{align*}
    p(x,r,\tau,A) = 
    \frac{1}{2} \left\{ 1 - \sgn(x) \zeta(x,r,\tau) \right\} .
\end{align*}
$$

Does this actually make a cline?
```{r show_cline}
pcline <- function (x,r,tau=1,sigma=1) {
    (1/2)*( 1 - sign(x)*zeta(x=x/sqrt(tau*sigma^2),r=r*tau) )
}
xx <- seq(-5,5,length.out=500)
plot(xx, pcline(xx,r=1), type='l', xlab="distance from cline center", ylab="prob of inherited from the left" )
rr <- seq(0,2,length.out=11)
pp <- sapply( rr, function (r) { pcline(xx,r=r) } )
matplot(xx,pp,type='l',lty=1,col=rainbow(length(rr)),add=TRUE)
legend("topright",lty=1,col=rainbow(length(rr)),legend=paste("r=",rr))
```
